\\textsc{Mathematical Description}

### 2022-03-18 100-math-description/GR-sslcm-math-description.Rmd
Change all table tables from XXX_table to XXX-table
bookdown doesn't like underscores in cross reference names for some reason

The model is framed in the Bayesian mode of inference, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain population dynamics parameters to take on only biologically plausible values through the use of prior information -- we intentionally employ nearly exclusively uninformative prior information, but in several rare cases we were required to bring in auxiliary information in the form of reasonably strong priors (Table \ref{tab:param_table}).
A final benefit of using the Bayesian framework is the seamless propagation of uncertainty from directly estimated parameters to derived quantities (of which there are many) -- all unknown quantities, either estimated parameters or deterministic functions of estimated parameters, receive a posterior distribution in the Bayesian framework.
Due to the analytical intractability of the complexity of the model, we employ Markov Chain Monte Carlo methods (implemented via the JAGS probabilistic programming language) to sample from the joint posterior distribution of all unknown quantities in the model.

We begin our description of the mathematical structure of the life cycle model with an overview of several syntax and statistical conventions we employ.
In our mathematical representation of quantities in the model, we have striven for consistency, but this has created somewhat non-intuitive conventions which require devoted explanation.
All quantities are defined (i.e., symbology and the specific equations that use them) in Tables \ref{tab:index_table}, \ref{tab:symbol_table}, \ref{tab:param_table}, and \ref{tab:data_table} for reference and a catalog of all equations is provided in Table \ref{tab:equation_table}.

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
\label{eq:generic-random}
\end{equation}
where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:
\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma4by4}
\end{equation}
That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:
\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma2by2}
\end{equation}
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \ref{tab:param_table}) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \ref{eq:generic-Sigma4by4} and \ref{eq:generic-Sigma2by2}.

@@

We further discovered that not all maturity and ocean survival terms were uniquely identifiable based solely on adult return abundance and age composition -- we thus relied on reasonably strong priors to inform $\dot{\phi}^{O^{1} \rightarrow O^{2}}_{o=\mathrm{NOR},j}$ and $\dot{\phi}^{O^{2} \rightarrow O^{3}}_{o=\mathrm{NOR},j}$ (Table \ref{tab:param_table}).

@@

Now, we must place the maturing portion of each of $O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$ into the correct year and age of return to the mouth of the Columbia River ($R^{b}_{y,k,o,j}$).
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age at maturity in year $y + k_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $k_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3).
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:
\begin{equation}
\begin{aligned}
R^{b}_{y+k_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
\label{eq:get-Rbs}
\end{equation}
Note, however, that this leaves 12 age/year combinations per origin unpopulated (the first 3 missing years for age-3 returns, first 4 missing years for age-4 returns, and first 5 missing years for age-5 returns).
This is because the life cycle dynamics must start somewhere, and no juvenile process model outcomes existed that would ultimately become these $R^{b}_{y,k,o,j}$ values in these early years.
Thus, to initialize the adult returns for natural-origin fish (and as a result, the rest of the life cycle), we estimated these 12 year/age return abundances as free parameters with fairly restrictive priors (Table \ref{tab:param_table}) with boundaries loosely informed by the ranges of adult returns at age observed in the early years of the data time series.
Initial abundance of hatchery-origin returns was handled by the "straying model", described below.

@@

No data sources were used to inform ocean juvenile population dynamics -- estimation of these quantities was enabled by (_i_) reasonably precise information about the abundance and composition of fish entering the ocean and returning to natal tributaries, (_ii_) the simplifying assumptions described above (i.e., time-constant second and third year ocean survival and perfectly correlated but offset origin-specific survival), and (_iii_) through the use of reasonably strong priors for second and third year ocean survival (Table \ref{tab:param_table}).

@@

```{r index_table}
# read in csv version of table
tab = read.csv("01-indices.csv", allowEscapes = TRUE)

@@

```{r symbol_tab_cap}
symbol_tab_cap = "The symbology used to represent the key states (i.e., abundance at life stage/group) and rates (i.e., transition probabilities among states) in the presentation of the state-space life cycle model for Grande Ronde spring Chinook salmon. The majority of rates presented in this table are hierarchically structured/estimated, where the population- and/or origin-specific expected value and standard deviation are free parameters (or a function of free parameters) presented in Table \\ref{tab:param_table}. Equation numbers reference any equation that uses that quantity -- these may include the equation where the quantity is first defined, where it is used in process model, or where it is used in the observation model."
```

```{r symbol_table}
tab = read.csv("02-symbols.csv", allowEscapes = TRUE)

@@

```{r param_table}
tab = read.csv("03-params.csv", allowEscapes = TRUE)

@@

```{r data_table}
tab = read.csv("04-data.csv", allowEscapes = TRUE)

@@

```{r equation_table}
tab = read.csv("05-equations.csv", allowEscapes = TRUE)

>>>>

The model is framed in the Bayesian mode of inference, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain population dynamics parameters to take on only biologically plausible values through the use of prior information -- we intentionally employ nearly exclusively uninformative prior information, but in several rare cases we were required to bring in auxiliary information in the form of reasonably strong priors (Table \ref{tab:param-table}).
A final benefit of using the Bayesian framework is the seamless propagation of uncertainty from directly estimated parameters to derived quantities (of which there are many) -- all unknown quantities, either estimated parameters or deterministic functions of estimated parameters, receive a posterior distribution in the Bayesian framework.
Due to the analytical intractability of the complexity of the model, we employ Markov Chain Monte Carlo methods (implemented via the JAGS probabilistic programming language) to sample from the joint posterior distribution of all unknown quantities in the model.

We begin our description of the mathematical structure of the life cycle model with an overview of several syntax and statistical conventions we employ.
In our mathematical representation of quantities in the model, we have striven for consistency, but this has created somewhat non-intuitive conventions which require devoted explanation.
All quantities are defined (i.e., symbology and the specific equations that use them) in Tables \ref{tab:index-table}, \ref{tab:symbol-table}, \ref{tab:param-table}, and \ref{tab:data-table} for reference and a catalog of all equations is provided in Table \ref{tab:equation-table}.

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
\label{eq:generic-random}
\end{equation}
where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:
\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma4by4}
\end{equation}
That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:
\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma2by2}
\end{equation}
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \ref{tab:param-table}) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \ref{eq:generic-Sigma4by4} and \ref{eq:generic-Sigma2by2}.

@@

We further discovered that not all maturity and ocean survival terms were uniquely identifiable based solely on adult return abundance and age composition -- we thus relied on reasonably strong priors to inform $\dot{\phi}^{O^{1} \rightarrow O^{2}}_{o=\mathrm{NOR},j}$ and $\dot{\phi}^{O^{2} \rightarrow O^{3}}_{o=\mathrm{NOR},j}$ (Table \ref{tab:param-table}).

@@

Now, we must place the maturing portion of each of $O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$ into the correct year and age of return to the mouth of the Columbia River ($R^{b}_{y,k,o,j}$).
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age at maturity in year $y + k_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $k_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3).
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:
\begin{equation}
\begin{aligned}
R^{b}_{y+k_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
\label{eq:get-Rbs}
\end{equation}
Note, however, that this leaves 12 age/year combinations per origin unpopulated (the first 3 missing years for age-3 returns, first 4 missing years for age-4 returns, and first 5 missing years for age-5 returns).
This is because the life cycle dynamics must start somewhere, and no juvenile process model outcomes existed that would ultimately become these $R^{b}_{y,k,o,j}$ values in these early years.
Thus, to initialize the adult returns for natural-origin fish (and as a result, the rest of the life cycle), we estimated these 12 year/age return abundances as free parameters with fairly restrictive priors (Table \ref{tab:param-table}) with boundaries loosely informed by the ranges of adult returns at age observed in the early years of the data time series.
Initial abundance of hatchery-origin returns was handled by the "straying model", described below.

@@

No data sources were used to inform ocean juvenile population dynamics -- estimation of these quantities was enabled by (_i_) reasonably precise information about the abundance and composition of fish entering the ocean and returning to natal tributaries, (_ii_) the simplifying assumptions described above (i.e., time-constant second and third year ocean survival and perfectly correlated but offset origin-specific survival), and (_iii_) through the use of reasonably strong priors for second and third year ocean survival (Table \ref{tab:param-table}).

@@

```{r index-table}
# read in csv version of table
tab = read.csv("01-indices.csv", allowEscapes = TRUE)

@@

```{r symbol_tab_cap}
symbol_tab_cap = "The symbology used to represent the key states (i.e., abundance at life stage/group) and rates (i.e., transition probabilities among states) in the presentation of the state-space life cycle model for Grande Ronde spring Chinook salmon. The majority of rates presented in this table are hierarchically structured/estimated, where the population- and/or origin-specific expected value and standard deviation are free parameters (or a function of free parameters) presented in Table \\ref{tab:param-table}. Equation numbers reference any equation that uses that quantity -- these may include the equation where the quantity is first defined, where it is used in process model, or where it is used in the observation model."
```

```{r symbol-table}
tab = read.csv("02-symbols.csv", allowEscapes = TRUE)

@@

```{r param-table}
tab = read.csv("03-params.csv", allowEscapes = TRUE)

@@

```{r data-table}
tab = read.csv("04-data.csv", allowEscapes = TRUE)

@@

```{r equation-table}
tab = read.csv("05-equations.csv", allowEscapes = TRUE)
